Unifying the communicable disease spreading paradigm with Gompertzian growth

### 2022-12-09 ResponseToReviewers.txt

Answer: I see that the the word "instantaneous" is inaccurate in this context and have changed this paragraph to 
"The SIRD model above also considers average macroscopic behavior of an ensemble of microscopic units justified through mean-field theory \cite{smilkov2014beyond} which does not consider network effects explicitly. Rather, all entities are connected and communicating *through long-range interactions* as shown by Mombach et al. (2002) \cite{mombach2002mean}.
"

The idea was to quote the extra assumptions that are present in mean field theories which do not have a physical basis in the communicable disease paradigm, namely that pathogens have long-range interactions, which by construction would imply near-luminal effect-velocities, which is a clear violation of this paradigm.

p.6 …the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.
Comment: The authors should explain: what is the nature and origin of this globally affecting biosphere stressor, which plays a key role in the interpretation of the results obtained. The context of the work suggests that it plays a similar role to the cosmological constant in the general relativity, guaranteeing the stability (stationarity) of the solutions.

Answer: Here we have intentionally left out a discussion of such stressor mechanisms and causes due to the uncertainty of such speculations. We do not wish to derail an otherwise scientific investigation and in some ways falsification of the communicable disease paradigm into somewhat speculative alternative theories. However, after some deliberation, we have added a few paragraphs on mechanitic and observational evidence for an environmental stressor theory. Although this mixing of biological mechanisms and strictly physical modelling might be spreading ourselves too thin, so to speak, so we have keept the former discussion to a minimum and relied mostly on the existing literature.

### 2022-12-09 arxiv/main.tex
expand on mechanistic and evidence discussion

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,liu2020emerging,medica2020covid,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to a Coronavirus infection, so why did these spikes all occur within the same time frame? While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model. This alternative theory represents an alternative paradigm in disease spreading dynamics. We will then present a novel bridge between these two paradigms which connects the logistic growth model with the Gompertzian growth model as an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible}; a bridge built on higher order interaction terms.

@@

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text. Although the Gompertz curve can also be obtained with a simple linear fit using Eq. \ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with the basic reproduction number defined as $\mathcal{R}_0=\gamma\xi/\kappa$ \citep{carletti2020covid}. The $95^{th}$ percent confidence interval is shown in brackets and computed with the parametric bootstrap assuming a multivariate normal distribution on the parameters and 10000 generated bootstrap samples \citep{efron2012bayesian}. Fitting is done with Python-Scipy and a 6-day moving average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}

\section*{An alternative theory for observed mortality}
\label{seq:alt}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability. [PLACEHOLDER: This stressor could be in the form of anomalous geomagnetic storms, or man-made disturbances in the atmosphere. For example, the ionosphere, which is the reservoir of the earth's electromagnetic potential, could have been disturbed either by man-made or natural events, causing organisms to experience an immune response. The exact mechanism of such hypotheses would need further study.]

@@

We can now further appreciate the Richards parametrization as a transition from non-collaborative to maximally collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by Petroni et al. (2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq. \ref{eq:rypdal} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$. 
Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at near-luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected. We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth.

@@

The basic reproduction number is modeled as the product of $\eta \xi / \kappa$ in Eq. \ref{eq:SIRD_D}, and its associated variance is obtained through a parametric bootstrap assuming a multivariate normal distribution with mean and covariance equal to the estimate from the non-linear least squares technique. The $95^{th}$ percent confidence interval is obtained after 10000 parameter samples drawn from this distribution, each computing a resampled estimate of $\mathcal{R}_0$ \citep{efron2012bayesian}.

@@

% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\citep{risken1996fokker},
% \begin{equation}
% P(z, t|z', t') = \sqrt{\frac{\beta}{2\pi \sigma (1 - e^{-2\beta (t-t')})}}\exp{\left[ -\frac{\beta (z - e^{-\beta(t-t')}z')^2}{2\sigma(1-e^{-2 \beta(t-t')})} \right]}.
% \end{equation}

@@

\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \citep{dong2020interactive}.}

>>>>

In addition to the characteristic patterns of these national mortality curves, other observations that conflict more generally with the communicable disease models are 1) the lack of correlation between population density and mortality (or infection) rates \cite{Hamidi2020,Hamidi2020a,Carozzi2020,liu2020emerging,medica2020covid,khavarian2021high,barak2021urban}, 2) no clear association between inter-generational relationships and fatality rates \cite{Arpino2020}, and 3) the widespread morbidity and mortality spikes seen during spring 2020 among birds \cite{Fischer2021}, dogs \cite{PeriseBarrios2021}, rabbits \cite{Duff2020,Hu2020,Fukui2021}, elephants \cite{Aarde2021}, and horses \cite{Kambayashi2021,King2020,Lu2020,CastilloOlivares2020}. Many of these animal mortality spikes are not attributable to a Coronavirus infection, so why did these spikes all occur within the same time frame? While these observations are not reconcilable within the current communicable disease paradigm, we present here an alternative theory that can reconcile all observations; a theory based on an environmental stressor with fewer parameters than the most basic communicable disease model.

This environmental stressor could be in the form of short-term solar and geomagnetic disturbances, which have been associated with physiological changes in cardiac system processes \cite{vieira2022geomagnetic,Alabdulgader2018,ghione1998geomagnetic,chernouss2001geophysical,stoupel1995ambulatory,stoupel2011twenty,malin1979correlation,stoupel1993sudden}, melatonin/serotonin balance \cite{burch1999geomagnetic,rapoport1997effects,bergiannaki1996seasonal}, endothelial activation and inflammation \cite{schiff2022role}, and calcium fluxes \cite{blackman1990elf}, all with the ability to cause neurological and pulmonary adverse effects. More general effects have also been studied in this context, e.g. overall mortality, suicide, metal disorders, homicides, and epileptic seizures \cite{zilli2019geomagnetic,gordon2003effect,kay1994geomagnetic,kay2004schizophrenia,halberg2005chronomics,berk2006ambient,doronin1998effect,persinger1995sudden}. Cherry (2002) put forward a mechanism whereby the ionosphere acts as a coupling agent between geomagnetic disturbances and physiological effects based on the disturbance of the background electromagnetic oscillations between the surface of the Earth and the lower ionosphere, called the Schumann Resonance signal \cite{cherry2002schumann,schumann1952strahlungslosen}. The Schumann Resonance would then couple with the brain to alter calcium fluxes and the melatonin/serotonin balance, which was a causal link observed by Wever (1974) through the circadian rhythm \cite{wever1974elf}.

While the Schumann Resonance mechanism is postulated to be caused by natural geomagnetic disturbances, others have looked at possible causative links through man-made radiation stressors and have found a similar pattern in the symptomology of a Coronavirus infection and radiation injury \cite{rios2021commonalities,Rubik2021}.

Whatever the cause, the environmental stressor theory represents an alternative paradigm in disease spreading dynamics. Our goal with this study is to present a novel bridge between these two paradigms which at its core is a connection between the logistic growth model and the Gompertzian growth model. This connection represents an alternative to the parametrization given by Richard (1959) \cite{richards1959flexible} and is instead a bridge built on higher order interaction terms between the susceptible and the infected populations. In doing this we shall show from a physical argument that the Gompertz model is incompatible with disease spreading through social interactions.

@@

\insertPdfFig{Gompertz_vs_SIRD_lnln_infty}{Cumulative number of deceased from the Coronavirus, transformed by $g(Y)=\ln{\ln{({Y_{\infty} / Y(t))}}}$ plotted against number of days elapsed after $Y(t) / Y_{max}>0.005$, comparing an SIRD model (green) with a Gompertz model (yellow) for a variety of countries in the period Jan-May 2020. Both models are fit using non-linear least squares according to equations \ref{eq:GompertzODE} and \ref{eq:SIRD_D} in the text. Although the Gompertz curve can also be obtained with a simple linear fit using Eq. \ref{eq:GompertzODE2param}, it is here obtained using non-linear least squares to put both models on equal footing. The temporal evolution of the SIRD model is obtained from the Runge-Kutta algorithm, while the Gompertz has a closed form for its temporal evolution. Each plot is annotated with the basic reproduction number defined as $\mathcal{R}_0=\gamma\xi/\kappa$ \cite{carletti2020covid}. The $95^{th}$ percent confidence interval is shown in brackets and computed with the parametric bootstrap assuming a multivariate normal distribution on the parameters and 10000 generated bootstrap samples \cite{efron2012bayesian}. Fitting is done with Python-Scipy and a 6-day moving average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}

\section*{An alternative theory for observed mortality}
\label{seq:alt}
An alternative line of reasoning that does not rely on the framework of communicable diseases is that the biosphere was perturbed by an instantaneous external stressor, initiating a stress response to eventually bring mortality rates back to stability.

@@

We can now further appreciate the Richards parametrization as a transition from non-collaborative to maximally collaborative growth as $\theta \rightarrow 0$. 
This feature of $\theta$ was also obtained by Petroni et al. (2020) \cite{petroni2020logistic} by interpreting the $\theta-$logistic growth rate in Eq. \ref{eq:rypdal} as non-linear resource availability dependent on the overall abundance, with linearity reached at $\theta=1$. The maximally coherent scenario implies that all units in the network are exposed to the stressor simultaneously.
Molski and Konarski (2003) \cite{molski2003coherent} made a similar observation that Gompertzian growth is the coherent state in a quantum mechanical system with a time-dependent potential, an interpretation which sheds further light on the temporal nature of the postulated stress response.
This quantum mechanical system has also been used to describe coherent energy states of diatomic molecules in space \cite{morse1929diatomic}. 
In the field of quantum physics, \emph{coherence} is a well-defined mathematical property first explored by Glauber (1963) \cite{glauber1963coherent} in the context of electromagnetic fields. The fact that we observe the Gompertz curve in both the microscopic quantum scales and the macroscopic national scales is noteworthy, and suggests that both systems share commonalities and means of communication.

Observing Gompertzian growth through the communicable diseases paradigm, the perceived pathogen would have to travel at luminal speeds throughout the population, exemplified by the immediate decrease in the relative growth rate of the infected. We are thus based on physical grounds forced to reject the concept of a pathogen in the context of Gompertzian growth.

@@

The basic reproduction number is modeled as the product of $\eta \xi / \kappa$ in Eq. \ref{eq:SIRD_D}, and its associated variance is obtained through a parametric bootstrap assuming a multivariate normal distribution with mean and covariance equal to the estimate from the non-linear least squares technique. The $95^{th}$ percent confidence interval is obtained after 10000 parameter samples drawn from this distribution, each computing a resampled estimate of $\mathcal{R}_0$ \cite{efron2012bayesian}.

@@

% To be explicit, if we consider the notation from Section \ref{seq:alt} and let $P(z, t|z', t')$ be the transition probability from some growth and time coordinate $z', t'$ to some other growth and time coordinate $z, t$, than the transition probability of the Olstein-Uhlenbeck process in \ref{eq:microscopic} is given by\cite{risken1996fokker},
% \begin{equation}
% P(z, t|z', t') = \sqrt{\frac{\beta}{2\pi \sigma (1 - e^{-2\beta (t-t')})}}\exp{\left[ -\frac{\beta (z - e^{-\beta(t-t')}z')^2}{2\sigma(1-e^{-2 \beta(t-t')})} \right]}.
% \end{equation}

@@

\insertPdfFig{Gompertz_vs_SIRD}{Cumulative diseased plotted against number of days after $Y(t) / Y_{max}>0.005$, comparing an SIRD with a Gompertz model. Both models are fit using non-linear least squares according to equations (\ref{eq:GompertzODE}) and (\ref{eq:SIRD_D}) in the text. Fitting is done with Python-Scipy and a 6-day windowed average of deaths. Observations are taken from the Github repository compiled by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University, Baltimore, USA \cite{dong2020interactive}.}
